---
title: "f2PilotBE"
output:
  github_document:
    toc: true
    toc_depth: 3
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


The `f2PilotBE` package is designed to calculate the similarity factor *f*~2~ for C~max~ and AUC, as an alternative analysis approaches for the assessment of bioavailability/bioequivalence (BA/BE) studies.

This package is based on the publications of [Henriques *et al.* (2023)](https://doi.org/10.3390/pharmaceutics15051430). In this work, the authors proposed the use of the geometric mean (G~mean~) *f*~2~ factor using a cut-off of 35, for the comparison of the absorption rate (given by the maximum observed concentration [C~max~]) of Test and Reference formulations. For the tested simulated scenarios, G~mean~ *f*~2~ factor using a cut-off of 35 showed a good relationship between avoiding type I and type II errors.


## The Similarity f2 Factor

The similarity *f*~2~ factor is a mathematical index widely used to compare dissolution profiles, evaluating their similarity, using the percentage of drug dissolved per unit of time. 

The similarity *f*~2~ factor, proposed by Moore and Flanner in 1996, is derived from the mean squared difference, and can be calculated as a function of the reciprocal of mean squared-root transformation of the sum of square differences at all points:

\deqn{
{f_2} = 50 \cdot \log \left( 100 \cdot \left[ 1 + \frac{1}{n} \sum_{t=1}^{t=n} \left(\overline{R}_t - \overline{T}_t\right)^2 \right]^{-0.5} \right)
}

$f_{2}=50\cdot\ log \biggl(100\cdot \biggl[1+\frac{1}{n} \sum_{t=1}^{t=n}{(\bar{R}_{t} - \bar{T}_{t})^2} \biggr]^{-0.5} \biggr)$,

where $f_{2}$ is the similarity factor, $n$ is the number of time points, and $\bar{R}_{t}$ and $\bar{T}_{t}$ are the mean percentage of drug dissolved at time $t$, for Reference and Test products respectively.

The *f*~2~ similarity factor ranges from 0 (when $\bar{R}_{t} - \bar{T}_{t} = 100 \%$, at all $t$) to 100 (when $\bar{R}_{t} - \bar{T}_{t} = 0 \%$, at all $t$).

The figure below presents the distribution of *f*~2~ similarity factor as a function of mean difference. *f*~2~ similarity factor is derived from the mean squared difference and can be calculated as a function of the reciprocal of the mean squared-root transformation of the sum of square differences at all points. An average difference of <span style="color: red;">10%</span>, <span style="color: green;">15%</span>, and <span style="color: blue;">20%</span> from all measured time points results in a *f*~2~ value of 50 (<span style="color: red;">red</span> dotted lines), 41 (<span style="color: green;">green</span> dotted lines) and 35 (<span style="color: blue;">blue</span> dotted lines), respectively.

```{r, echo=FALSE, results = 'asis', fig.cap="Distribution of ƒ2 similarity factor as a function of mean difference. ƒ2 similarity factor is derived from the mean squared difference and can be calculated as a function of the reciprocal of the mean squared-root transformation of the sum of square differences at all points. An average difference of 10%, 15%, and 20% from all measured time points results in a ƒ2 value of 50 (red dotted lines), 41 (green dotted lines) and 35 (blue dotted lines), respectively."}
image = "https://pub.mdpi-res.com/pharmaceutics/pharmaceutics-15-01430/article_deploy/html/images/pharmaceutics-15-01430-g002.png?1683533911"
cat(paste0('<center><img src="', image,  '"></center>')) 
```

(Figure from [[Henriques *et al.* (2023)](https://doi.org/10.3390/pharmaceutics15051430)])

As previously proposed by the authors, the concept of similarity factor *f*~2~ can be applied as an alternative to the average bioequivalence analysis [[Henriques *et al.* (2023)](https://doi.org/10.3390/pharmaceutics15051430)]. The similarity between Test and Reference products by means of *f*~2~ was evaluated through the comparison of arithmetic (A~mean~) and geometric (G~mean~) means of plasma concentration-time profiles derived from the simulated individual pharmacokinetic profiles.

### For Cmax

*f*~2~ can be used to assess the similarity on the rate of drug absorption by normalizing Test and Reference mean concentration-time profiles to the maximum plasma concentration (C~max~) derived from the mean Reference profile, until Reference C~max~ is observed (Reference t~max~) [[Henriques *et al.* (2023)](https://doi.org/10.3390/pharmaceutics15051430)]:

$C_{t}^{N}=100\cdot\frac{\bar{C}_{t}}{C_{max,R}}$, where $0 \le t \le t_{max,R}$,

where $C_{t}^{N}$ is the normalized concentration at time $t$, $\bar{C}_{t}$ is the mean (Test or Reference) concentration at time $t$, $C_{max,R}$ is the C~max~ of the Reference mean concentration-time profile, and $t_{max,R}$ the time of observation of $C_{max,R}$. The similarity *f*~2~ factor is calculated as

$C_{max}f_{2}=50\cdot\ log \biggl(100\cdot \biggl[1+\frac{1}{n} \sum_{t=1}^{t=n}{(R_{t}^{N} - T_{t}^{N})^2} \biggr]^{-0.5} \biggr)$,

where $n$ is the number of time points until Reference $t_{max}$, and $R_{t}^{N}$ and $T_{t}^{N}$ are the normalized concentration at time $t$, for Reference and Test products respectively.


## Installation

### From GitHub

To install the development version of `f2PilotBE`, start by installing the devtools package from CRAN:

```{r, echo=TRUE, eval=FALSE}
install.packages("devtools")
```

Then the development version of `f2PilotBE` can be installed from GitHub as:

```{r, echo=TRUE, eval=FALSE}
library(devtools)
install_github("LittlePetinga/f2PilotBE")
```


## Examples

### Load Package

Following installation, load the `f2PilotBE` package and the `ggplot2` package.

```{r, echo=TRUE, eval=FALSE}
# Load packages
library("ggplot2")      # required for plotting
library("f2PilotBE")
```


### Geometric Mean Concentration

Import individual concentration data and calculate the G~mean~ concentration-time profiles for each Treatment.

```{r, echo=TRUE, eval=FALSE}
# Import individual concentration data
dta_id <- read.csv('dta_id.csv')


# Calculate the geometric mean for each Treatment, by time point
dta <- data.frame(t(tapply(dta_id$Concentration, 
                           dta_id[,c('Treatment','Time')], 
                           geomean)))
dta$Time <- as.numeric(row.names(dta))
row.names(dta) <- c()

```


### Calculate f2 Factor for Cmax

Calculate the *f*~2~ for C~max~ from the G~mean~ concentration-time profiles:

```{r, echo=TRUE, eval=FALSE}
# Calculate f2 Factor for Cmax
# when treatment data is pivoted:
f2.Cmax(dta, Time = 'Time', Reference = 'R', Test = 'T',
        Trt.cols = TRUE, details = FALSE, plot = TRUE)
```

In the example above, treatment data is pivoted. can also be calculated when treatment data is stacked:

```{r, echo=TRUE, eval=FALSE}
# To stack treatment information
dta <- cbind(dta[3],stack(dta[1:2]))
names(dta)[names(dta) %in% c('values','ind')] <- c('Concentration', 'Treatment')

# Calculate f2 Factor for Cmax
# when treatment data is pivoted:
f2.Cmax(dta, Time = 'Time', Concentration = 'Concentration', 
        Treatment = 'Treatment', Reference = 'R', Test = 'T',
        Trt.cols = FALSE)
```


## References

Henriques, S.C.; Albuquerque, J.; Paixão, P.; Almeida, L.; Silva, N.E. (2023). Alternative Analysis Approaches for the Assessment of Pilot Bioavailability/Bioequivalence Studies. *Pharmaceutics*. *15*(5), 1430. [10.3390/pharmaceutics15051430](https://doi.org/10.3390/pharmaceutics15051430)

Henriques, S.C.; Paixão, P.; Almeida, L.; Silva, N.E. (2023). Predictive Potential of C~max~ Bioequivalence in Pilot Bioavailability/Bioequivalence Studies, through the Alternative *f*~2~ Similarity Factor Method. *Pharmaceutics*. *15*(10), 2498. [10.3390/pharmaceutics15102498](https://doi.org/10.3390/pharmaceutics15102498).

Moore, J.W.; Flanner, H.H. (1996). Mathematical Comparison of Curves with an Emphasis on in Vitro Dissolution Profiles. *Pharm. Technol.*. *20*, 64--74.
